National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-17 Manuscript.Rmd
revisions to manuscript, worked out practical effect sizes

---
title             : "National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs"
shorttitle        : "Beliefs and Wellbeing"
author: 
  - name          : "Ben Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

@@

abstract: |
 Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and reliance on cross-sectional samples cast doubts. Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective well-being. 
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "4778"
bibliography      : ["b-references.bib"]

@@

```{r runanalysis, include = FALSE}
source('short.R')
```
The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics raise three challenges.

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this features as a "tautology" in spirituality/mental health research [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures dressed in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr].  However previous research does not converge on a unique set of meanings. Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg]. However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being [@King2013-cg]. Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangle the concept of spirituality from concepts of religion.

A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples. To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out [@Garssen2016-km].  However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. To address the key questions in spirituality psychological health research, then, requires longitudinal investigations.

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of N = 21,705 (N = 20,979 with complete responses).

@@

## Statistical Analyses

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.  We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. 
The model equation(s) are described in Equation 1.

## Equation 1

\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Ids}
\end{aligned}
\]

@@

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with  Id as random effects. The model included Id as random effects.  The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001). Within this model:

  - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458).  We infer that Skeptics personal well-being is not increasing in the Skeptic population
  - The effect of Beliefs_SpiritExcludesGod_is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics. 
  - The effect of BeliefsGodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p < .001). We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics. 
  - The effect of BeliefsGodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people have greater personal well-beingwellbeing.
  - The effect of Deprivation.S is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people have lower personal well-beingwellbeing.
  - The effect of Edu.S is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people greater personal well-beingwellbeing
  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). Employed people have greater personal well-beingwellbeing.
  - The effect of EthnicCatsMaori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
  - The effect of EthnicCatsPacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).
  - The effect of EthnicCatsAsian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
  - The effect of Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).
  - The effect of Having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
  - The effect of Pol.Orient.S is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).
  - The effect of Urban is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05).

@@

  - The effect of Years X SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p < .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
  - The effect of Years X GodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
  - The effect of Years X GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813).  Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.

@@

  - The effect of Years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
  - The effect of BeliefsSpiritExcludesGod  is reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics. 
  - The effect of BeliefsGodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
  - The effect of BeliefsGodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction. 
  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
  - The effect of Deprivation.S is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
  - The effect of Edu.S is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
  - The effect of Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
  - The effect of EthnicCatsMaori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
  - The effect of EthnicCatsPacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p < .05).  People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
  - The effect of EthnicCatsAsian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
  - The effect of Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
  - The effect of HasPartner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
  - The effect of Pol.Orient.S is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
  - The effect of Urban is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).

@@

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USElifsat_expected-1.png")
```
  - The interaction of Years X Beliefs_SpiritExcludesGod_ is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
  - The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799).  Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.

## Imputed data analysis

To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample. This analysis is describe in Supplement 1. Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction. Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction. Figures 3 and Figures 4 describe the predicted slopes of change. The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.  In all analyses those who believe in a spirit or life-force are expected to experience greater growth in well-being over time.

Figure 3 about here

@@

Figure 4 about here

@@

Figure 5 about here

@@

The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples. Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.

We find that \dots
1. Lower subjective well-being among spirit beliefs population at outset of study, surpassed (LS) or surpassing (PWI) by the end of the interval
2. Holds after adjusting for demograph
3. Holds for Belief in God

Questions:
1. Why initially lower well-being in NZ? 
2. Why the slope for spirit beliefs (and not God)?

Importance:
1. Methods: Disentangling
2. Growing population: practical importance

These findings demonstrate the importance of investigating spiritual beliefs as a domain of interest in its own right. In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr]. Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg]. As such, it is important to understand the ramifications of belief in a spirit for psychological health. 
To assess the relationship between a belief in a spirit or life force and subjective well-being, we improve upon past methods in three ways.

>>>>

---
title             : "National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs"
shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
author: 
  - name          : "Ben Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

@@

abstract: |

  Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being. 
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "5015"
bibliography      : ["b-references.bib"]

@@

```{r runanalysis, include = FALSE}
source('short.R')
```
The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three credible challenges.

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this approach as "tautological" [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr]. Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings. Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg]. As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being [@King2013-cg]. Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out [@Garssen2016-km]. Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. However, To address the key questions in spirituality psychological health research requires longitudinal investigations. Unfortunately the relevant datasets are rare.

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of $N$ = 21,705 ($N$ = 20,979 with complete responses).

@@

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.  We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. 
The model equation(s) are described in Equation 1.

## Equation 1
\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
\end{aligned}
\]

@@

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, withId as random effects. The model included Id as random effects.  The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001). Within this model:

  - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458).  We infer that Skeptics personal well-being is not increasing in the Skeptic population
  - The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics. 
  - The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p < .001). We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics. 
  - The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people have greater personal well-beingwellbeing.
  - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people have lower personal well-beingwellbeing.
  - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people greater personal well-beingwellbeing
  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). Employed people have greater personal well-beingwellbeing.
  - The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
  - The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).
  - The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
  - The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).
  - The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
  - The effect of Political Conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).
  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05).

@@

  - The effect of years $\times$ belief in a SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p < .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
  - The effect of years $\times$ in GodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
  - The effect of Years $\times$ belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813).  Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.

@@

  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
  - The effect of beliefs in SpiritExcludesGodis reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics. 
  - The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
  - The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction. 
  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
  - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
  - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
  - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
  - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
  - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p < .05).  People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
  - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
  - The effect of identifying as Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
  - The effect of having a Parnter is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
  - The effect of Political Conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).

@@

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USElifsat_expected-1.png")
```
  - The interaction of years $\times$ beliefs in SpiritExcludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of years $\times$ beliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
  - The effect of years $\times$ beliefs in GodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799).  Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.

## Imputed data analysis
To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample. This analysis is describe in Supplement 1. Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction. Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction. Figures 3 and Figures 4 describe the predicted slopes of change. The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave. In all analyses those who believe in a spirit or life-force are expected to experience greater growth in well-being over time.

Figure 3 about here:

@@

Figure 4 about here:

@@

Figure 5 about here:

@@

The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples. Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over 9 waves (8 years) in nationally diverse western populations who differ in the dimensions of spiritual and religious beliefs. We observe that:

1. At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05). (See Figure 2)

Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02). (See Figure 3)
